Towards open collaboration in biomedical and health research: An open collaborative scoping review

### 2025-02-10 doc/paper.qmd

This is a scoping review, so we followed the framework described in
[@Arksey2005, @Levac2010] as well as the guidelines outlined in the
PRISMA-ScR statement [@Tricco2018].

@@

While we didn't change our search terms for the initial search and
extraction, we found that our search terms were not precise enough and
got many irrelevant scholarly works. The full R code to filter down the
search results is kept in the
[`R/exlusions.R`](https://github.com/science-collective/scoping-review/blob/main/R/exclusions.R)
file on the GitHub repository. For example, we extracted a large number
of scholarly works describing surgery such as "open wound", electronics
such as "open circuit", or environmental such as "open water" that we
had to post-process and exclude.

@@

Across the papers found in our literature search, X out of 11 discuss
actual experiences on implementing open collaboration, albeit in
different organizational levels. These range from lab level (Turoman et
al) center level (Bush et al and Grange et al), multi-center level
(ManyPrimates et al and Grange et al) and field level: psychology
(Alessandorni et al), Spinal Cord injury (Torres-Espin et al) and
neuroimaging (Niso et al). All of these share in common the need for an
infrastructure that promotes open science and open collaboration via
starting building a community of collaborators or making tools (usually
online resources) that researchers can use to reach FAIR (Finndable,
Accessible, Interoperable and Reusable) goals while doing research.
Before delving in those, here we provide the specific needs for
implementing open collaboration in each organizational level, as well as
describing the differences in approaches among experiences between and
across levels (Figure 2).

@@

For center-level only one paper was identified (Bush et al), providing
the example of a center of neuroimaging describing first how they
identified the needs for open collaboration, then the individual-level
practices that researchers should adapt and then the center-level
changes to promote reaching FAIR goals.

@@

Interestingly, the level of organization with most papers is the
multi-center level with two papers from ManyPrimates et al (ManyPrimates
2019 and ManyPrimates 2021) and Grange et al. In this organization level
the most urgent needs are: 1) identifying the needs for open
collaboration and 2) creating an infrastructure that allows those needs
to be satisfied. Indeed, Grange et al describes a systematic review as
an exploration of the needs for 11 UKRN centers to implement open
collaborative practives. In this paper they specially focus on tools and
resources among the 11 institutions and make clear the need for starting
to utilize open resources as a first step to adapt open collaborative
practices. The first paper from ManyPrimates et al \[REF\] describes the
start of another multi-center collaboration, albeit from different
institutions. It is interesting to observe that in this situation, the
complex issue is to start the collaboration itself, which was ignited
thanks to a symposium and an email chain among future collaborators. In
contrast, ManyPrimates could quickly develop an open collaborative
structure with relative ease in comparison to changing an already solid
infrastructure, like in the case of Grange et al.

>>>>

This is a scoping review, so we followed the framework described in
[@Arksey2005; @Levac2010] as well as the guidelines outlined in the
PRISMA-ScR statement [@Tricco2018].

@@

While we didn't change our search terms for the initial search and
extraction, we found that our search terms were not precise enough and
got many irrelevant scholarly works. The full R code to filter down the
search results is kept in the
[`R/exclusions.R`](https://github.com/science-collective/scoping-review/blob/main/R/exclusions.R)
file on the GitHub repository. For example, we extracted a large number
of scholarly works describing surgery such as "open wound", electronics
such as "open circuit", or environmental such as "open water" that we
had to post-process and exclude.

@@

Across the papers found in our literature search, X out of 11 discuss
actual experiences on implementing open collaboration, albeit in
different organizational levels. These range from lab level
[@Turoman2022], center level [@Bush2022; @Network2023], multi-center
level [@ManyPrimates2021; @Network2023] and field level: psychology
[@Alessandroni2022], spinal cord injury [@TorresEspin2021] and
neuroimaging [@Niso2022]. All of these share in common the need for an
infrastructure that promotes open science and open collaboration via
starting building a community of collaborators or making tools (usually
online resources) that researchers can use to reach FAIR (Finndable,
Accessible, Interoperable and Reusable) goals while doing research.
Before delving in those, here we provide the specific needs for
implementing open collaboration in each organizational level, as well as
describing the differences in approaches among experiences between and
across levels (Figure 2).

@@

For center-level only one paper was identified [@Bush2022], providing
the example of a center of neuroimaging describing first how they
identified the needs for open collaboration, then the individual-level
practices that researchers should adapt and then the center-level
changes to promote reaching FAIR goals.

@@

Interestingly, the level of organization with most papers is the
multi-center level with two papers from ManyPrimates [@ManyPrimates2021;
@Primates2019] and one from the Reproducibility Network [@Network2023].
In this organization level the most urgent needs are: 1) identifying the
needs for open collaboration and 2) creating an infrastructure that
allows those needs to be satisfied. Indeed, the Reproducibility Network
[@Network2023] describes a systematic review as an exploration of the
needs for 11 UKRN centers to implement open collaborative practives. In
this paper they specially focus on tools and resources among the 11
institutions and make clear the need for starting to utilize open
resources as a first step to adapt open collaborative practices. The
first paper from ManyPrimates [@Primates2019] describes the start of
another multi-center collaboration, albeit from different institutions.
It is interesting to observe that in this situation, the complex issue
is to start the collaboration itself, which was ignited thanks to a
symposium and an email chain among future collaborators. In contrast,
ManyPrimates could quickly develop an open collaborative structure with
relative ease in comparison to changing an already solid infrastructure,
like in the case of @Network2023.
